Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2023-08-15 docs/Manuscript/Manuscript.Rmd
continue transcribing straightforward responses to reviewers

---
title: "Clinical prediction using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University at Florida, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which an (often conventional) model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.

@@

## Protocol

Our methods followed prepared plans, which we collect into a protocol in the Appendix.
We also note deviations from these plans and the reasons for them.

### PRISMA checklist

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms.

@@

Following the selection sources aligning with our review, we collected characteristics of included studies (see the Appendix). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, evaluation/performance/comparison measures (& their values), name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r fig:prisma, fig.align='center', out.width="80%", fig.cap = "\\label{fig:prisma}PRISMA-S flow chart.", echo=FALSE}
knitr::include_graphics("../../fig/fig-prisma.pdf")
```

@@

## Overview

The 25 included studies were analyzed based on 20 characteristics (Table \ref{tab:synthesis}), and we describe some observations here. The studies were published in a variety of journals, with some of greater frequency, though no single journal published exceptionally many. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, and _Hindawi - Journal of Healthcare Engineering_ published 3. Another distinguishing factor was the years of publication of the included studies (Figure \ref{fig:year}). While they stretch back to the late 1990s, they have become more common in recent years, which suggests that this is still an active, though not rapidly expanding methodological approach. We note that, while CBR in health and medicine originated as early as 1990, these studies did emerge early on once CBR had established itself. So the idea has been "in the air" for as long as CBR has been in use.

```{r tab:synthesis, echo=FALSE, message=FALSE, warning=FALSE}
googlesheets4::read_sheet(
  ss = "1tpWMhYH2pyRT55K7n2J2XFs-kEV_JTuCDmXzJ4BBgNo", sheet = 1L
) |>
  rename_with(snakecase::to_snake_case) |>
  filter(is.na(synthesis)) |> 
  arrange(date_of_publication) |> 
  transmute(
    Citation = paste0("@", zotero_key),
    Date = format(date_of_publication, "%Y %b %d"),
    Task = objective,
    Aim = ifelse(generalizable_knowledge == "Science", "Knowledge", "Practice"),
    Source = type_of_data,
    Type = range_of_data,
    `Cases` = number_of_cases_incidents,
    `Features` = number_of_predictors_features
  ) |>
  mutate(Task = ifelse(Task == "Recommendation", "Decision Support", Task)) |> 
  mutate(across(c(Cases, Features), sapply, FUN = format, big.mark = ",")) |>
  mutate(across(c(Cases, Features), sapply, FUN = paste0, collapse = "; ")) |>
  # mutate(across(c(Cases, Features), str_split, pattern = "; ")) |> 
  # mutate(Dimensions = purrr::map2_chr(
  #   Cases, Features,
  #   ~ str_c(.x, .y, sep = " × ", collapse = "; ")
  # )) |> 
  mutate(across(
    everything(),
    # ~ stringr::str_replace_all(., "<U+001B>", "")
    textclean::replace_non_ascii
  )) ->
  tab_synthesis
tab_synthesis |> 
  select(-Date) |> 
  # prevent numeric list columns from dominating the horizontal space
  mutate(across(
    c(-Citation, -Cases, -Features),
    ~ str_c(., "\\hspace{6em}")
  )) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:synthesis}",
      "Studies included in the method synthesis,",
      " arranged by the earliest the study is known to have been public."
    )
  )
```

```{r fig:year, fig.align='center', out.width="80%", fig.cap = "\\label{fig:year}Number of publications in our sample each year.", echo=FALSE}
knitr::include_graphics("../../fig/fig-years.png")
```

@@

```{r fig:framework, fig.align='center', out.width="80%", fig.cap = "\\label{fig:framework}General framework for localized models.", echo=FALSE}
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

# Data and code availability

@@

## Protocol

### Research question

@@

### Purpose of review

@@

### Search

@@

### Screening process

### Selection process

### Analysis and synthesis

## Search strings

@@

>>>>

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
author:
  - name: Adi Cohen
    email: ac4569@my.nsu.edu
    affiliation: 1
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: 3
    correspondingauthor: true
address:
  - code: 1
    address: College of Osteopathic Medicine, Nova Southeastern University, 3200 S University Drive, Davie, FL 33328
  - code: 2
    address: Smathers Libraries Academic Research & Consulting Services, University at Florida, P. O. Box 117000, Gainesville, FL 32611-7000
  - code: 3
    address: Laboratory for Systems Medicine, University of Florida, P.O. Box 100225 JHMHC, Gainesville, FL 32610-0225
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) to clinical and health tasks have given rise to a family of techniques in which an (often conventional) model is fitted to a subset of labeled cases, selected for relevance or similarity to an unlabeled case to which the model is applied. These localized models, as we propose to call them, may perform as well as "black-box" models while retaining direct interpretations. Their adoption has been hindered by fragmented development and implementation challenges. Our goal in this scoping review was to describe the clinical and health applications of localized models to date and to derive a general framework for their design and evaluation.

@@

Our methods followed prepared plans, which we describe in detail in the Appendix.
We also note deviations from these plans and the reasons for them.

## PRISMA checklist

@@

We derived the eligibility criteria itemized in the Introduction from a seed set of previously read studies. Based on these criteria, we formulated search strings for five platforms: PubMed, Web of Science, Academic Search Premier, Google Scholar, and MathSciNet. We finalized the PubMed search first, then adapted it to the other platforms (Sections \ref{sec:appendix-search}).

@@

Following the selection sources aligning with our review, we collected characteristics of included studies (Section \ref{sec:availability}). The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, evaluation/performance/comparison measures (& their values), name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

```{r fig:prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:prisma}
#| PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s)."
knitr::include_graphics("../../fig/fig-prisma.pdf")
```

@@

## Bibliographic and methodological properties

The 25 included studies were analyzed based on 20 characteristics (see Section \ref{sec:availability}), and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, and _Hindawi - Journal of Healthcare Engineering_ published 3. Another interesting pattern lay in the years of publication of the included studies (Figure \ref{fig:year}). While they stretch back to the late 1990s, they have become more common, which suggests that this is still an active, though not rapidly expanding, methodological approach. We note that, while CBR in health and medicine originated as early as 1990, these studies did emerge early on once CBR had established itself. So the idea has been "in the air" for as long as CBR has been in use.

```{r tab:synthesis, echo=FALSE, message=FALSE, warning=FALSE}
tab_synthesis <- read_rds(here::here("tab/synthesis.rds"))
tab_synthesis |> 
  select(-Date) |> 
  # prevent numeric list columns from dominating the horizontal space
  mutate(across(
    c(-Citation, -Cases, -Features),
    ~ str_c(., "\\hspace{6em}")
  )) |> 
  knitr::kable(
    align = "lllllrr",
    caption = paste0(
      "\\label{tab:synthesis}",
      "Studies included in the method synthesis,",
      " arranged by the earliest the study is known to have been public."
    )
  )
```

```{r fig:year, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:year}
#| Number of publications in our sample each year."
knitr::include_graphics("../../fig/fig-years.png")
```

@@

```{r fig:framework, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "\\label{fig:framework}
#| General framework for localized models."
knitr::include_graphics("../../fig/fig-typology.pdf")
```

@@

# Data and code availability
\label{sec:availability}

@@

## Research question

@@

## Purpose of review
\label{sec:appendix-purpose}

@@

## Search
\label{sec:appendix-search}

@@

### Search strings

@@

## Screening process
\label{sec:appendix-screening}

## Selection process
\label{sec:appendix-selection}

## Analysis and synthesis
